---
title: "Interrupted Time Series ARIMA Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
    code_folding: hide 
    
output_dir: Output
---

```{r setup, include=FALSE}

library(tidyverse)
library(lubridate)
library(tsibble)
library(fable)
library(feasts)
library(forecast)
library(scales)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
) 

#importing functions
source("functions.r")
``` 


# Introduction

This document uses an interrupted time series ARIMA analysis to evaluate changes in mortality (assault and all-cause) following the June 2021 Bill 8 policy intervention restricting access to abortion in Texas. 

Data are from the CDC Multiple Cause of Death Database, including monthly Texas resident death counts from 2018-Dec 2023. We fit both female and male, though note that we plan to just report female, including one male model as a counterfactual comparison. 

# Methods 

R code is adapted from this [RPubs article on tidy structure for ITS ARIMAs](https://rpubs.com/jcorbin/tidy_itt_arima)  

The ARIMA model incorporates:

- **Autoregressive terms (AR)** to capture dependence on previous observations

- **Integrated differencing (I)** to remove non-stationary trends

- **Moving average terms (MA)** to account for serial correlation in residual errors

The intervention is modeled using:

- **Step variable**: immediate level change after the intervention

- **Ramp variable**: gradual slope change after the intervention

Counterfactual forecasts are generated using the **pre-intervention ARIMA structure**, representing expected mortality trajectories if the policy had not occurred.   

*Note*: We also conducted a sensitivity analysis where instead of using a step and ramp variable to indicate whether the interruption has passed, we subset the data to Pre-Interruption period, fit an arima, then fit those arima specifications to the full dataset. 

## Analysis Structure 

1. All-Cause 

We will start by reviewing the ARIMA for **All-Cause Mortality**, so we may capture any changes in death trends overall, accounting for the fact that deaths by assault are often misrecorded by government agencies. 

2. Reproductive-aged Assault Mortality  

We will then explore deaths by assault among women aged 15-44.

3. All-age Assault Mortality

Lastly, we will investigate assault mortality among all ages, not just reproductive.  






# All-Cause Mortality

## Results 

### Male 

```{r}

monthly_m.ts <- prepare_ts("../CleanData/all_cause_repro_m.csv")
monthly_m.ts <- create_intervention(monthly_m.ts, "2021-06-01")

m_arima <- fit_arima(monthly_m.ts)

fc_m <- counterfactual_forecast(monthly_m.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_m,
  monthly_m.ts,
  "2021-06-01",
  "Observed vs Counterfactual Male All-Cause Deaths"
)
```

### Female 

```{r}

monthly_f.ts <- prepare_ts("../CleanData/all_cause_repro_f.csv")
monthly_f.ts <- create_intervention(monthly_f.ts, "2021-06-01")

f_arima <- fit_arima(monthly_f.ts)

fc_f <- counterfactual_forecast(monthly_f.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_f,
  monthly_f.ts,
  "2021-06-01",
  "Observed vs Counterfactual Female All-Cause Deaths"
)


``` 

## Model Estimates 

The **step coefficient** represents the immediate change in deaths following the intervention, while the **ramp coefficient** estimates the monthly change in slope after the intervention.

```{r}
report(m_arima)
report(f_arima)
``` 

## Model Diagnostics 

<details> <summary><b>Residual Diagnostics</b></summary> 
```{r}

m_arima %>% gg_tsresiduals()
f_arima %>% gg_tsresiduals()

``` 

</details> <details> <summary><b>Ljung-Box Test</b></summary> 
```{r}

augment(m_arima) %>% features(.resid, ljung_box)
augment(f_arima) %>% features(.resid, ljung_box)

```
</details> 

## Sensitivity Analysis 

<details> <summary><b>Pre-Intervention Model Selection - Female and Male</b></summary> 
```{r}

pre_data_f <- monthly_f.ts %>%
  filter(month_year < "2021-06-01")

pre_data_m <- monthly_m.ts %>%
  filter(month_year < "2021-06-01")

f_pre <- pre_data_f %>%
  model(arima = ARIMA(Deaths))

m_pre <- pre_data_m %>%
  model(arima = ARIMA(Deaths))

report(f_pre)
report(m_pre)

```
</details> 

<details> <summary><b>Fitting with Full Data</b></summary> 
```{r}

f_sensitivity <- monthly_f.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,0,1),
      stepwise = FALSE
    )
  ) 

m_sensitivity <- monthly_m.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,0) + PDQ(1,0,0),
      stepwise = FALSE
    )
  )

report(f_sensitivity)
report(m_sensitivity)
```
</details> 

<details> <summary><b>Forecasted Graphs</b></summary> 
```{r}
fc_counterfactual_f <- f_pre %>%
  forecast(h = 28+4)

fc_counterfactual_m <- m_pre %>%
  forecast(h = 28+4)

autoplot(fc_counterfactual_f, monthly_f.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date("2021-06-01")),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Female Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal() 

autoplot(fc_counterfactual_m, monthly_m.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date("2021-06-01")),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Male Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal()
```
</details>  

### Explicit seasonal + trend terms 

As pointed out by Brooke, we may want to explicitly model the seasonal and trend terms in the ARIMA itself. 

We will first make month a factor: 
```{r}
monthly.ts <- monthly.ts %>%
  mutate(
    month_factor = factor(lubridate::month(month_year))
  ) 

m_sensitivity <- monthly.ts %>% 
  model(
    arima = ARIMA(Deaths ~ step + ramp + t + month_factor,
                  stepwise = FALSE)
  )

report(m_sensitivity) 

#it'd be nice here if I made a function to plot the counterfactual predicted trend?
``` 

t is the global linear trend term (t= time), and month_factor forces recurring seasonal patterns to persist. 
Other option: use Fourier terms (number of sine/cosine pairs)
- we don't know the seasonality this explicitly, might not be the way to go??


# Assault Mortality 

## Results 

### Male 
```{r}

monthly_m.ts <- prepare_ts("../CleanData/monthly_m.csv")
monthly_m.ts <- create_intervention(monthly_m.ts, "2021-06-01")

m_arima <- fit_arima(monthly_m.ts)

fc_m <- counterfactual_forecast(monthly_m.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_m,
  monthly_m.ts,
  "2021-06-01",
  "Observed vs Counterfactual Male Assault Deaths"
)

```

### Female 
```{r}

monthly_f.ts <- prepare_ts("../CleanData/monthly_f.csv")
monthly_f.ts <- create_intervention(monthly_f.ts, "2021-06-01")

f_arima <- fit_arima(monthly_f.ts)

fc_f <- counterfactual_forecast(monthly_f.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_f,
  monthly_f.ts,
  "2021-06-01",
  "Observed vs Counterfactual Female Assault Deaths"
)

```

## Model Estimates 

```{r}

report(m_arima)
report(f_arima)

```

## Model Diagnostics 

<details> <summary><b>Residual Diagnostics</b></summary> 
```{r}

m_arima %>% gg_tsresiduals()
f_arima %>% gg_tsresiduals()

``` 

</details> <details> <summary><b>Ljung-Box Test</b></summary> 
```{r}

augment(m_arima) %>% features(.resid, ljung_box)
augment(f_arima) %>% features(.resid, ljung_box)

```
</details>  

## Sensitivity Analysis 

<details> <summary><b>Pre-Intervention Model Selection - Female and Male</b></summary> 
```{r}

pre_data_f <- monthly_f.ts %>%
  filter(month_year < "2021-06-01")

pre_data_m <- monthly_m.ts %>%
  filter(month_year < "2021-06-01")

f_pre <- pre_data_f %>%
  model(arima = ARIMA(Deaths))

m_pre <- pre_data_m %>%
  model(arima = ARIMA(Deaths))

report(f_pre)
report(m_pre)

```
</details> 

<details> <summary><b>Fitting with Full Data</b></summary> 
```{r}

f_sensitivity <- monthly_f.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,0,1),
      stepwise = FALSE
    )
  ) 

m_sensitivity <- monthly_m.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,0) + PDQ(1,0,0),
      stepwise = FALSE
    )
  )

report(f_sensitivity)
report(m_sensitivity)
```
</details> 

<details> <summary><b>Forecasted Graphs</b></summary> 
```{r}
fc_counterfactual_f <- f_pre %>%
  forecast(h = 28+4)

fc_counterfactual_m <- m_pre %>%
  forecast(h = 28+4)

autoplot(fc_counterfactual_f, monthly_f.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date("2021-06-01")),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Female Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal() 

autoplot(fc_counterfactual_m, monthly_m.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date("2021-06-01")),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Male Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal()
```
</details> 

## All-Age Assault Mortality 

## Results 

### Male 
```{r}

monthly_m.ts <- prepare_ts("../CleanData/all_age_m_assault.csv")
monthly_m.ts <- create_intervention(monthly_m.ts, "2021-06-01")

m_arima <- fit_arima(monthly_m.ts)

fc_m <- counterfactual_forecast(monthly_m.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_m,
  monthly_m.ts,
  "2021-06-01",
  "Observed vs Counterfactual All-Age Male Assault Deaths"
)

``` 

### Female 

```{r}

monthly_f.ts <- prepare_ts("../CleanData/all_age_f_assault.csv")
monthly_f.ts <- create_intervention(monthly_f.ts, "2021-06-01")

f_arima <- fit_arima(monthly_f.ts)

fc_f <- counterfactual_forecast(monthly_f.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_f,
  monthly_f.ts,
  "2021-06-01",
  "Observed vs Counterfactual All-Age Female Assault Deaths"
)

```

## Model Diagnostics 

<details> <summary><b>Residual Diagnostics</b></summary> 
```{r}

m_arima %>% gg_tsresiduals()
f_arima %>% gg_tsresiduals()

``` 

</details> <details> <summary><b>Ljung-Box Test</b></summary> 
```{r}

augment(m_arima) %>% features(.resid, ljung_box)
augment(f_arima) %>% features(.resid, ljung_box)

```
</details> 

## Sensitivity Analysis 

<details> <summary><b>Pre-Intervention Model Selection - Female and Male</b></summary> 
```{r}

pre_data_f <- monthly_f.ts %>%
  filter(month_year < "2021-06-01")

pre_data_m <- monthly_m.ts %>%
  filter(month_year < "2021-06-01")

f_pre <- pre_data_f %>%
  model(arima = ARIMA(Deaths))

m_pre <- pre_data_m %>%
  model(arima = ARIMA(Deaths))

report(f_pre)
report(m_pre)

```
</details> 

<details> <summary><b>Fitting with Full Data</b></summary> 
```{r}

f_sensitivity <- monthly_f.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,0,1),
      stepwise = FALSE
    )
  ) 

m_sensitivity <- monthly_m.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,0) + PDQ(1,0,0),
      stepwise = FALSE
    )
  )

report(f_sensitivity)
report(m_sensitivity)
```
</details> 

<details> <summary><b>Forecasted Graphs</b></summary> 
```{r}
fc_counterfactual_f <- f_pre %>%
  forecast(h = 28+4)

fc_counterfactual_m <- m_pre %>%
  forecast(h = 28+4)

autoplot(fc_counterfactual_f, monthly_f.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date("2021-06-01")),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Female Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal() 

autoplot(fc_counterfactual_m, monthly_m.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date("2021-06-01")),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Male Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0,0)) +
  theme_minimal()
```
</details>  



