---
title: "Cleaning"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
    code_folding: hide  
    
output_dir: Output
---

```{r setup}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(broom)
library(viridis)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

#importing functions
source("functions.r")
``` 

# Introduction 

This script contains the code used to clean the raw datasets accessed from CDC Wonder contained within the `/RawData` directory. All datasets generated from this script will be saved in the `/CleanData` directory.  

Functions used to clean the datasets may be accessed from the `/Scripts/functions.R` file.  


# Female Monthly Assault Counts 

```{r}
rawdata_female <- read_excel("../RawData/MonthlyAssaultsFemaleRepro.xlsx") 
monthly_f <- rawdata_female %>% 
  mutate(month_year = my(Month),
         Deaths = na_if(as.character(Deaths), "Suppressed"),
         Deaths = as.numeric(Deaths)) %>% 
   select(-Notes,-Population,-"Crude Rate", -Month, -"Month Code", -"Crude Rate Lower 95% Confidence Interval", -"Crude Rate Upper 95% Confidence Interval") 
head(monthly_f) 

``` 

We need to apply the same cleaning methodology, this time for Texas resident males in the same age group (15-44). 

# Male Monthly Assault Counts 

```{r}
rawdata_male <- read_excel("../RawData/MaleMonthlyAssaults.xlsx") 
monthly_m <- rawdata_male %>% 
  mutate(month_year = my(Month)) %>% 
   select(-Notes,-Population,-"Crude Rate", -Month, -"Month Code") 
head(monthly_m) 


monthly_m %>% 
  slice(35:55)
               

``` 
It should be noted we are missing death data for females for May 2019 (2019-05-01) and September 2019 (2019-09-01). We need to fill in this hole, so we will investigate filling this gap in the data by querying both genders and subtracting the deaths for males to reveal how many were among females.

```{r}
rawdata_both_genders <- read_excel("../RawData/2019_monthly_both_sexes.xlsx") 
both_genders_2019 <- rawdata_both_genders %>% 
  mutate(month_year = my(Month)) %>% 
   select(-Notes,-Population,-"Crude Rate", -Month, -"Month Code") 
head(both_genders_2019)  

#selecting the dates we need 

both_genders_2019 %>% 
  filter(month_year %in% c("2019-05-01", "2019-09-01")) 

#107 in may, 96 in sept 

monthly_m %>% 
  filter(month_year %in% c("2019-05-01", "2019-09-01"))  
#98 in may, 89 in sept 

#may: 107-98 = 9
#sept: 96-89 = 7

#dropping the problematic rows for females
monthly_f <- monthly_f %>%
  filter(
    !(month_year %in% as.Date(c("2019-05-01", "2019-09-01")))
  )

monthly_f <- monthly_f %>%
  bind_rows(
    tibble(
      month_year = as.Date(c("2019-05-01", "2019-09-01")),
      Deaths     = c(11, 7),
      sex        = "Female"
    )
  )

```

Next we are going to merge the datasets to create an all sex version.

```{r}

monthly_all <- bind_rows( #combining datasets
  monthly_f %>% mutate(sex = "Female"),
  monthly_m %>% mutate(sex = "Male")
)

monthly_m %>% #checking which months are missing in the female df
  anti_join(monthly_f, by = "month_year") %>%
  select(month_year)

monthly_all <- monthly_all %>%
  group_by(sex) %>% 
  arrange(month_year) %>%
  mutate(time = row_number())
range(monthly_all$time)

monthly_f <- monthly_f %>%
  arrange(month_year) %>%
  mutate(time = row_number())
range(monthly_f$time)

monthly_m <- monthly_m %>%
  arrange(month_year) %>%
  mutate(time = row_number())

range(monthly_m$time)

#looks correct! Saving the datasets

write_csv(monthly_all, "../CleanData/monthly_all.csv")
write_csv(monthly_m, "../CleanData/monthly_m.csv")
write_csv(monthly_f, "../CleanData/monthly_f.csv")
``` 

# All Cause Female Reproductive age 

Next we are going to clean this dataset so we may examine how all-cause deaths may differ from assault only. This will be important as deaths by assault are underreported and often misclassified. Let's start by importing the data and applying the cleaning function for .csv's.

```{r} 

all_repro_f <- read.csv("../RawData/texas_allcause_female1544.csv")

all_repro_f <- clean.csv(all_repro_f)

head(all_repro_f) 

``` 
Let's check out the data: 
```{r}
ggplot(all_repro_f, aes(month_year, Deaths))+ 
  geom_line()

``` 

Looks good! Let's save the all cause female data to our cleaned folder. 
```{r}
write_csv(all_repro_f, "../CleanData/all_cause_repro_f.csv")
``` 

Males this time:
```{r} 

all_repro_m <- read.csv("../RawData/texas_allcause_male1544.csv") 
glimpse(all_repro_m)

all_repro_m <- all_repro_m %>% 
  mutate(month_year = my(Month)) %>% 
  select(-Month, -Month.Code, -Notes,-Population:-Crude.Rate.Upper.95..Confidence.Interval) %>% 
  filter(month_year < "2024-01-01") %>% 
  mutate( time = row_number ())


ggplot(all_repro_m, aes(month_year, Deaths))+ 
  geom_line() 

write_csv(all_repro_m, "../CleanData/all_cause_repro_m.csv")

```

# Assault all age female 

Next we are going to clean the data for all ages female, cod by assault. 

```{r}
all_age_f_assault <- read_excel("../RawData/All_Ages_Female.xlsx")
all_age_f_assault <- clean.xcel(all_age_f_assault) 
head(all_age_f_assault) 

all_age_m_assault <- read.csv("../RawData/all_age_assault_male.csv")
all_age_m_assault <- all_age_m_assault %>% 
   mutate(month_year = my(Month)) %>% 
  select(-Month, -Month.Code, -Notes,-Population:-Crude.Rate.Upper.95..Confidence.Interval) %>% 
  filter(month_year < "2024-01-01") %>% 
  mutate( time = row_number ()) 

write_csv(all_age_m_assault, "../CleanData/all_age_m_assault.csv")

  
``` 
Same as before, let's briefly check out what the data looks like: 
```{r}
ggplot(all_age_f_assault, aes(month_year, Deaths))+ 
  geom_line()
```  

Let's write our csv: 
```{r}
write_csv(all_age_f_assault, "../CleanData/all_age_f_assault.csv")
```

# Age 15+ 

Let's also check out the data for texas deaths by assault, this time just with ages 15+. 

```{r}
assault_15_plus <- read_excel("../RawData/assault_15_plus.xlsx") 
assault_15_plus <- clean.xcel(assault_15_plus) 
head(assault_15_plus) 

ggplot(assault_15_plus, aes(month_year, Deaths))+ 
  geom_line() 

write_csv(assault_15_plus, "../CleanData/assault_15_plus")
```


# National control set 

This time we are going to pull ages 15-44 deaths by assault nationally. 

```{r}
nat_f_assault <- read_excel("../RawData/national_f_assault.xlsx") 
nat_f_assault <- clean.xcel(nat_f_assault) 
head(nat_f_assault) 

ggplot(nat_f_assault, aes(month_year, Deaths))+ 
  geom_line() 

write_csv(nat_f_assault, "../CleanData/nat_f_assault.csv")
``` 

# Disaggregated Race 

We pulled the deaths for assault and all-cause, but we are just going to focus on all-cause to increase power. 

```{r}
h <- read.csv("../RawData/RacialDisaggregation/All_cause/H.csv")
nh <- read.csv("../RawData/RacialDisaggregation/All_cause/NH.csv") 


racial <- clean_racial(nh,h) 
write_csv(racial, "../CleanData/racial_all_cause.csv")
  
``` 


# COVID Deaths 

We need to clean the data with deaths attributed to COVID so we may create an all-cause minus COVID mortality dataset. 

Starting with pulling up the COVID mortality data (female reproductive age): 

```{r}
covid <- read.csv("../RawData/COVID/COVID.csv")
#checking import
glimpse(covid)
``` 

We have quite a few days where the death value is suppressed. 

```{r}
covid %>% 
  filter(Deaths == "Suppressed") %>% 
  select(Month)
```

```{r}
#Need to change suppressed values to be 0s and make deaths numeric. 
covid <- covid %>% 
    mutate(Deaths = as.numeric(na_if(Deaths, "Suppressed")),
          month_year = my(Month)) %>% 
  select(-Notes,-Month, -"Month.Code", -Population:-"Crude.Rate.Upper.95..Confidence.Interval") %>% 
  filter(month_year < "2024-01-01") %>% 
  mutate(time = row_number ())

``` 

Next, we will create new COD variables so when we delete the deaths from all cause we can match up accordingly: 

```{r}
covid <- covid %>% 
  mutate(cod = "covid")

all_repro_f <- all_repro_f %>% 
  mutate(cod = "all_cause") 

#checking if dates are the same: 
identical(
  sort(unique(covid$month_year)),
  sort(unique(all_repro_f$month_year))
)

#finding dates missing 
setdiff(covid$month_year, all_repro_f$month_year)
setdiff(all_repro_f$month_year, covid$month_year)

combined <- all_repro_f %>%
  left_join(covid, 
            by = c("month_year"),
            suffix = c("_all", "_covid"))

#subtracting deaths, making NA deaths 0
all_cause_minus_covid <- combined %>%
  mutate(
    Deaths = Deaths_all - coalesce(Deaths_covid, 0)
  ) %>%
  select(month_year, Deaths)

head(all_cause_minus_covid)

#checking that we have every month! 
full_dates <- seq.Date(
  from = as.Date("2018-01-01"),
  to   = as.Date("2023-12-01"),
  by   = "month"
)

missing_dates <- setdiff(full_dates, all_cause_minus_covid$month_year)

missing_dates

``` 

After creating our all cause minus covid, it looks as though we have successfully subtracted all and preserved all the dates in our study period. We can now save this cleaned dataset. 

```{r}

write_csv(all_cause_minus_covid, "../CleanData/all_cause_minus_covid.csv")

```

## Racial Disag 

```{r}
h_covid <- read.csv("../RawData/RacialDisaggregation/COVID/H.csv")
nh_covid <- read.csv("../RawData/RacialDisaggregation/COVID/NH.csv") 
head(h_covid)
head(nh_covid)

racial_covid <- clean_racial(nh_covid,h_covid)
head(racial_covid)

racial_covid <- racial_covid %>% 
  mutate(cod = "covid") %>% 
  filter(!race_eth %in% c("Native Hawaiian or Other Pacific Islander", "American Indian or Alaska Native"))  

h <- read.csv("../RawData/RacialDisaggregation/All_cause/H.csv")
nh <- read.csv("../RawData/RacialDisaggregation/All_cause/NH.csv") 
racial <- clean_racial(nh,h)
head(racial)

``` 

```{r}
#cleaning the two datasets: 
racial <- racial %>% 
  filter(!race_eth %in% c("Native Hawaiian or Other Pacific Islander", "American Indian or Alaska Native")) %>% 
  mutate(cod = "all_cause")  

head(racial_covid)
head(racial)

#checking if dates are the same: 
identical(
  sort(unique(racial_covid$month_year)),
  sort(unique(racial$month_year))
)


#combining
combined_racial <- racial %>%
  left_join(racial_covid, by = c("month_year", "race_eth"),
            suffix = c("_all", "_covid"))

#subtracting deaths, making NA deaths 0
racial_minus_covid <- combined_racial %>%
  mutate(
    Deaths = Deaths_all - coalesce(Deaths_covid, 0)
  ) %>%
  select(race_eth, month_year, Deaths)

head(racial_minus_covid)

#checking that we have every month! 
full_dates <- seq.Date(
  from = as.Date("2018-01-01"),
  to   = as.Date("2023-12-01"),
  by   = "month"
)

missing_dates <- setdiff(full_dates, racial_minus_covid$month_year)

missing_dates  

write_csv(racial_minus_covid, "../CleanData/racial_minus_covid.csv")
  
``` 

### Annual 

We are going to look at annual trends in terms of assault deaths for repro-age females, just in case there's something to be said there. 

```{r}
h_assault <- read.csv("../RawData/RacialDisaggregation/Assault/Annual/H.csv")
nh_assault <- read.csv("../RawData/RacialDisaggregation/Assault/Annual/NH.csv") 
head(h_assault)
head(nh_assault)

unique(nh_assault$Year)
unique(h_assault$Year)

racial_assault <- clean_racial_annual(nh_assault,h_assault)  


head(racial_assault) 

racial_assault %>% 
  group_by(race_eth) %>% 
  filter(is.na(deaths)) 

#missing too much info for Asian and American Indian/Alaska Native, excluding those from df 

racial_assault <- racial_assault %>% 
  filter(!race_eth %in% c("American Indian or Alaska Native", "Asian")) 

write.csv(racial_assault, "../CleanData/racial_assault_annual.csv")
```
