---
title: "All_Cause_Minus_COVID"

output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
    code_folding: hide 

---

```{r setup}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(broom)
library(viridis)
library(tsModel)
library(scales)
library(grid) 
library(tseries) 
library(tsibble) 
library(fable)
library(feasts) 
library(forecast)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
rmarkdown::render("All_Cause_Minus_COVID.Rmd", output_file = "Output/All_Cause_Minus_COVID.html")

source("functions.R")

```

This contains all code used to analyze the datasets found in the `/CleanData` directory, which were previously generated in the `/Scripts/Cleaning.rmd` file.

# Introduction

This analysis will replicate the May 2021 ARIMA, this time altering the all-cause death data to exclude deaths due to COVID, as they had a colinear increase over the same time period. 

# Descriptive Statistics 

We will start by comparing the all-cause deaths over the time period with the deaths just attributable to COVID. 

```{r all_cause_f}
all_cause <- read.csv("../CleanData/all_cause_repro_f.csv") 
all_cause.ts <- all_cause %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(all_cause.ts)

all_cause.ts|>
  gg_tsdisplay(y = Deaths, plot_type = "partial")
``` 

```{r covid_f}
covid <- read.csv("../")
```


# All-Cause Minus COVID

We will 




