---
title: "All Cause Minus COVID"

output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
    code_folding: hide 
    
output_dir: Output
---


```{r setup, include=FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(broom)
library(viridis)
library(tsModel)
library(scales)
library(grid) 
library(tseries) 
library(tsibble) 
library(fable)
library(feasts) 
library(forecast)
library(rmarkdown)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 

source("functions.R")
```

This contains all code used to analyze the datasets found in the `/CleanData` directory, which were previously generated in the `/Scripts/Cleaning.rmd` file.

# Introduction

This analysis will replicate the May 2021 ARIMA, this time altering the all-cause death data to exclude deaths due to COVID, as they had a colinear increase over the same time period. 

# Descriptive Statistics 

We will start by comparing the all-cause deaths over the time period with the all-cause deaths minus the ones attributable to covid. 

```{r all_cause_f}
all_cause <- read.csv("../CleanData/all_cause_repro_f.csv") 
all_cause.ts <- all_cause %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(all_cause.ts)

all_cause.ts|>
  gg_tsdisplay(y = Deaths, plot_type = "partial")
``` 

```{r covid_f}
all_minus_covid <- read.csv("../CleanData/all_cause_minus_covid.csv")

all_minus_covid.ts <- all_minus_covid %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year),
         time = row_number()) %>% 
  as_tsibble(index = month)
  
glimpse(all_minus_covid.ts)

all_minus_covid.ts |>
  gg_tsdisplay(y = Deaths, plot_type = "partial")
```

Looking at these graphs, it is pretty clear that once we remove COVID-attributable deaths, we see a very different picture: it looks like over the study period, all-cause deaths are generally increasing over time, with no observable outliers. 

We will continue with fitting an arima, though it is almost guaranteed that all the values forecasted will be completely expected. 

```{r, include = FALSE}
range(all_minus_covid.ts$time)
# Create variable representing step and ramp changes
intervention_month <- all_minus_covid.ts %>% 
  filter(month_year == "2021-05-01") %>% 
  pull(time)
 all_minus_covid.ts <-  all_minus_covid.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  )

#View the step and ramp variables
all_minus_covid.ts %>% 
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-05-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Visualizing Step and Ramp Variables",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) 
``` 

```{r, include = FALSE}
# Use automated algorithm to identify p/q parameters
# Specify first difference = 1 and seasonal difference = 1
all_minus_covid.arima <- all_minus_covid.ts %>% 
  model(arima = ARIMA(Deaths ~ step + ramp, stepwise = FALSE))

#Plotting model residuals
all_minus_covid.arima  %>% 
  gg_tsresiduals()

#Ljung-Box test
augment(all_minus_covid.arima) %>%  
  features(.resid, ljung_box)
``` 

```{r}
report(all_minus_covid.arima)

#Modeling counterfactual
counterfactual <- all_minus_covid.ts %>%
  filter(month_year < "2021-05-01") %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(counterfactual)

#forecast into post-intervention period 
fc_counterfactual <- counterfactual %>%
  forecast(h = 28+4) 

 autoplot(fc_counterfactual, all_minus_covid.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 5))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Female All-Cause Minus COVID Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```



