---
title: "Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true 
    
output_dir: Output
---
```{r setup}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(broom)
library(viridis)
library(tsModel)
library(scales)
library(grid) 
library(tseries) 
library(tsibble) 
library(fable)
library(feasts) 
library(forecast)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 

#importing functions
source("functions.r")

``` 

This contains all code used to analyze the datasets found in the `/CleanData` directory, which were previously generated in the `/Scripts/Cleaning.rmd` file. 

This document uses an interrupted time series ARIMA analysis to evaluate changes in mortality (assault and all-cause) following the Bill 8 policy intervention restricting access to abortion in Texas. Note that May 21, 2021 was when Bill 8 passed, though it later went into effect in September. Due to widespread media coverage in Texas, it may be reasonably assumed that reactions to the law may have began in June, lending to its designation as the interruption point for this study. May was not selected as only 10 days of the month remained after the bill got passed, biasing the results and potentially not providing enough time for information to disseminate. 

# Introduction

There are several studies that use the interrupted time series ARIMA, but we are taking inspiration from the methods described in [Singh et al., 2025](https://link.springer.com/article/10.1007/s00127-025-02902-7), a study which analyzes suicide deaths among reproductive-aged women post-*Dobbs*. 

This study also uses data from the Multiple Cause of Death database, and stratifies data by men and women by two age groups: 15-25 and 25-49. The outcomes are the monthly count of suicides nationally for reproductive-aged women overall and for the two age subgroups. The series of monthly suicides among men (per age group) were used as a control variable. The exposure was the timing of the *Dobbs* decision in June 2022. 

*Note: the following description is directly adapted from Singh et al. (2025). Future methods manuscripting will need to cite this source as well as adapt the phrasing.*

The ARIMA's autoregressive (AR) component should allow us to capture the relationship between current assault mortality counts and past observations. The moving average (MA) will model the residuals from previous time points (accounting for irregular fluctuations). The seasonality/trends that change over time will be addressed through differencing ("integrated"), which will remove secular trends and stabilize the mean. 

We may use the methods from Singh et al. (2025) in the following steps: 

1. Use the Box-Jenkins iterative pattern recognition routines to identify the ARIMA signature of the outcome in the pre-Bill 8 period (January, 2018 - August, 2021), controlling for the concomitant series of assault mortality among men of the same age group (15-44).  

2. Use the pre-Bill 8 ARIMA signature identified in step 1 to forecast (predict) values of the outcome for the 28-month period post-Bill 8 (i.e. September 2021 - December 2023). These predicted/fitted values will serve as counterfactuals and reflect the expected monthly count of female suicides if Bill 8 had not been passed. 

3. Use the ARIMA signature identified in step 1 and expected values from step 2 to obtain residual values (observed less fitted) of our outcome series. 

4. Use the pre-Bill 8 standard error of residuals to develop pre-Bill 8 95% confidence intervals (CI; prediction interval of monthly series of assault mortalities among Texas resident reproductive-aged women overall from January 2018 - August 2021). Extend the pre-Bill 8 prediction interval to the full residual series (January 2018 - December 2023). Next, note whether any residuals deviated from this prediction interval post-Bill 8 (i.e., September 2021 onward.)

5. If the outcome series exhibited outliers in the post-Bill 8 period following step 4, create a binary indicator of the month-specified timing of any identified outliers in residuals of assault mortalities among reproductive-aged women post-Bill 8 (1 for months with outliers, 0 otherwise) to quantify the average change in assault mortalities post-Bill 8. This will also help avoid multiple testing that could increase chances of a Type I error (false rejection of the null). 

6. Apply the binary-indicator of month-specific outliers (from step 5) to the ARIMA-derived residuals of outcomes to determine the magnitude of outcome deviation from expected levels in the months with residual outliers. 

2-tailed tests with significance level of p < 0.05 will be used. 


# Methods 

R code is adapted from this [RPubs article on tidy structure for ITS ARIMAs](https://rpubs.com/jcorbin/tidy_itt_arima)  

The ARIMA model incorporates:

- **Autoregressive terms (AR)** to capture dependence on previous observations

- **Integrated differencing (I)** to remove non-stationary trends

- **Moving average terms (MA)** to account for serial correlation in residual errors

The intervention is modeled using:

- **Step variable**: immediate level change after the intervention

- **Ramp variable**: gradual slope change after the intervention

Counterfactual forecasts are generated using the **pre-intervention ARIMA structure**, representing expected mortality trajectories if the policy had not occurred.   

*Note*: We also conducted a sensitivity analysis where instead of using a step and ramp variable to indicate whether the interruption has passed, we subset the data to Pre-Interruption period, fit an arima, then fit those arima specifications to the full dataset. We did not see any significant changes from our ARIMA fit, so we chose to proceed with reporting the results from the models chosen with the FABLE algorithm.

## Analysis Structure 

1. Reproductive-aged Assault Mortality  

We will start by exploring our primary research question of interest, which concerns how deaths by assault among women aged 15-44 changed post-Bill 8 announcement.  

2. All-age Assault Mortality

Next, we will investigate assault mortality among all ages, not just reproductive.  

3.  All-Cause Minus COVID

Lastly, we will  review the ARIMA for **All-Cause Mortality Minus COVID**, so we may capture any changes in death trends overall, accounting for the fact that deaths by assault are often misrecorded by government agencies. In our previous exploration, we discovered that the majority of deaths may be attributed to a spike in COVID cases in Aug/Sept 2021, so we are including this as a sensitivity analysis.


# Descriptive Statistics  

We are going to get some descriptive stats for our analyses of interest, using June as our interruption point. We are also going to explore racial differences over the years, acknowledging that due to data constraints we cannot describe these monthly (or fit to a model.)

We are going to start by exploring how many assault mortalities among Texas residents occured between the study period.  


# 1. Assault Mortality 

## Results  
### Female 
```{r}

monthly_f.ts <- prepare_ts("../CleanData/monthly_f.csv")
monthly_f.ts <- create_intervention(monthly_f.ts, "2021-06-01")

f_arima <- fit_arima(monthly_f.ts)

fc_f <- counterfactual_forecast(monthly_f.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_f,
  monthly_f.ts,
  "2021-06-01",
  "Observed vs Counterfactual Female Assault Deaths"
)

``` 

### Male 
```{r}

monthly_m.ts <- prepare_ts("../CleanData/monthly_m.csv")
monthly_m.ts <- create_intervention(monthly_m.ts, "2021-06-01")

m_arima <- fit_arima(monthly_m.ts)

fc_m <- counterfactual_forecast(monthly_m.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_m,
  monthly_m.ts,
  "2021-06-01",
  "Observed vs Counterfactual Male Assault Deaths"
)

``` 

## Model Estimates 

```{r}

report(f_arima)
report(m_arima)

``` 

## Model Diagnostics 

<details> <summary><b>Residual Diagnostics</b></summary> 
```{r}

m_arima %>% gg_tsresiduals()
f_arima %>% gg_tsresiduals()

``` 

</details> <details> <summary><b>Ljung-Box Test</b></summary> 
```{r}

augment(m_arima) %>% features(.resid, ljung_box)
augment(f_arima) %>% features(.resid, ljung_box)

```
</details>  

## 2. All-Age Assault Mortality 

## Results 

### Female 

```{r}

monthly_f.ts <- prepare_ts("../CleanData/all_age_f_assault.csv")
monthly_f.ts <- create_intervention(monthly_f.ts, "2021-06-01")

f_arima <- fit_arima(monthly_f.ts)

fc_f <- counterfactual_forecast(monthly_f.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_f,
  monthly_f.ts,
  "2021-06-01",
  "Observed vs Counterfactual All-Age Female Assault Deaths"
)

```

Just as a comparison (male): 
### Male 
```{r}

monthly_m.ts <- prepare_ts("../CleanData/all_age_m_assault.csv")
monthly_m.ts <- create_intervention(monthly_m.ts, "2021-06-01")

m_arima <- fit_arima(monthly_m.ts)

fc_m <- counterfactual_forecast(monthly_m.ts, "2021-06-01", 34)

plot_counterfactual(
  fc_m,
  monthly_m.ts,
  "2021-06-01",
  "Observed vs Counterfactual All-Age Male Assault Deaths"
)

``` 

## Model Diagnostics 

<details> <summary><b>Residual Diagnostics</b></summary> 
```{r}
m_arima %>% gg_tsresiduals()
f_arima %>% gg_tsresiduals()
``` 

</details> <details> <summary><b>Ljung-Box Test</b></summary> 
```{r}
augment(m_arima) %>% features(.resid, ljung_box)
augment(f_arima) %>% features(.resid, ljung_box)
```
</details> 

## 3. All-Cause Minus Covid 

```{r}
all_minus_covid <- prepare_ts("../CleanData/all_cause_minus_covid.csv") 
all_minus_covid <- all_minus_covid %>% 
  mutate(time = row_number())
all_minus_covid <- create_intervention(all_minus_covid, "2021-06-01")

m_covid_arima <- fit_arima(all_minus_covid)

fc_m_covid <- counterfactual_forecast(all_minus_covid, "2021-06-01", 34)

plot_counterfactual(
  fc_m_covid,
  all_minus_covid,
  "2021-06-01",
  "Observed vs Counterfactual Female All-cause Minus COVID Deaths"
)
```

## Model Estimates 

```{r}
report(m_covid_arima)
``` 

## Model Diagnostics 

<details> <summary><b>Residual Diagnostics</b></summary> 
```{r}
m_covid_arima %>% gg_tsresiduals()
``` 

</details> <details> <summary><b>Ljung-Box Test</b></summary> 
```{r}
augment(m_covid_arima) %>% features(.resid, ljung_box)
```
</details>  