---
title: "Exploration"
output: html_document
---

```{r setup}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(broom)
library(viridis)
library(tsModel)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```


This contains all code used to explore the datasets found in the `/CleanData` directory, which were previously generated in the `/Scripts/Cleaning.rmd` file. 

# Descriptive Analysis  

We are going to get some descriptive stats for our 3 analyses of interest, using June as our interruption point.

## 1. All-Cause 
```{r, include = FALSE}
#female all cause, reproductive age
f_all <- read.csv("../CleanData/all_cause_repro_f.csv")

# male all cause, reproductive age
m_all <- read.csv("../CleanData/all_cause_repro_m.csv")
 
#all cause MINUS covid (female only, repro age) 
f_minus_covid <- read.csv("../CleanData/all_cause_minus_covid.csv") 

#just covid, female, repro age

covid <- read.csv("../RawData/COVID/COVID.csv")
covid <- covid %>% 
    mutate(Deaths = as.numeric(na_if(Deaths, "Suppressed")),
          month_year = my(Month)) %>% 
  select(-Notes,-Month, -"Month.Code", -Population:-"Crude.Rate.Upper.95..Confidence.Interval") %>% 
  filter(month_year < "2024-01-01") %>% 
  mutate(time = row_number ())
```

**How many deaths occurred over the study period that were all cause? By sex?**

```{r}
f_all %>% 
  summarise(
    n_months      = n(),
    total_deaths  = sum(Deaths, na.rm = TRUE),
    monthly_mean  = mean(Deaths, na.rm = TRUE),
    monthly_sd    = sd(Deaths, na.rm = TRUE),
    .groups = "drop"
  )


m_all %>% 
  summarise(
    n_months      = n(),
    total_deaths  = sum(Deaths, na.rm = TRUE),
    monthly_mean  = mean(Deaths, na.rm = TRUE),
    monthly_sd    = sd(Deaths, na.rm = TRUE),
    .groups = "drop"
  )
```
Females had 33,290 over the study period. Males had 72,089 deaths. 105,379 total deaths occurred among reproductive aged-texas residents from January 2018 to Dec 2023. 

**What months had the most deaths for males and females? What were the top 3 death counts?** 

```{r}
f_all %>% 
  arrange(desc(Deaths)) %>% 
  select(month_year, Deaths) %>% 
  slice(1:3) 

m_all %>% 
  arrange(desc(Deaths)) %>% 
  select(month_year, Deaths) %>% 
  slice(1:3) 

```
For both males and females, the months with the highest death counts were August, September, and October of 2021. Deaths were higher among men than females for these months.

**How many female deaths were due to covid? For which months? What do they look like plotted against each other?** 

```{r}
#deaths due to covid
covid %>%  
  summarise(
    n_months      = n(),
    total_deaths  = sum(Deaths, na.rm = TRUE),
    monthly_mean  = mean(Deaths, na.rm = TRUE),
    monthly_sd    = sd(Deaths, na.rm = TRUE),
    .groups = "drop"
  )

#which months 
covid %>% 
  filter(!is.na(Deaths)& Deaths != 0)

#deaths without covid 
f_minus_covid %>%  
 summarise(
    n_months      = n(),
    total_deaths  = sum(Deaths, na.rm = TRUE),
    monthly_mean  = mean(Deaths, na.rm = TRUE),
    monthly_sd    = sd(Deaths, na.rm = TRUE),
    .groups = "drop"
  )

#plotting them both side by side
f_all <- f_all %>%
  mutate(month_year = as.Date(month_year))

covid <- covid %>%
  mutate(month_year = as.Date(month_year))

ggplot() +
  geom_line(data = f_all,
            aes(x = month_year, y = Deaths),
            color = "black") +
  geom_line(data = covid,
            aes(x = month_year, y = Deaths),
            color = "red") +
 scale_x_date(date_breaks = "2 months", date_labels = "%b %Y")+
  labs(x = "Month-Year",
       y = "Deaths") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

``` 

It's pretty clear here that a big portion of the death spike we see is from a covid outbreak in texas. 

**Let's plot and describe the deaths that are all causes, EXCEPT COVID deaths.** 
```{r}
#how many deaths over study period?
f_minus_covid %>%  
  summarize(n_deaths = sum(Deaths, na.rm = TRUE))

#largest death counts? 
f_minus_covid %>% 
  arrange(desc(Deaths)) %>% 
  select(month_year, Deaths) %>% 
  slice(1:3) 

#plotting 
f_minus_covid <- f_minus_covid %>% 
  mutate(month_year = as.Date(month_year))

vline_date <- as.Date("2021-05-01") 

ggplot(f_minus_covid)+
geom_line(data = f_minus_covid,
            aes(x = month_year, y = Deaths),
            color = "black") +
   geom_vline(xintercept = vline_date, linetype = "dashed", color = "red") +
  annotate("text",
           x = vline_date,
           y = 350, # slightly above top
           label = "June, 2021",
           angle = 90,
           vjust = -0.5, 
           color = "red")

``` 

It does not appear as though these results are unexpected. No outliers are likely. 

## 2. Reproductive-aged Assault Mortality  

Next, we will explore just reproductive-aged deaths by assault. 

```{r, include = FALSE}
monthly_f <- read.csv("../CleanData/monthly_f.csv") 
monthly_m <- read.csv("../CleanData/monthly_m.csv")
monthly_all <- read.csv("../CleanData/monthly_all.csv")

monthly_f <- monthly_f %>% 
  mutate(month_year = ymd(month_year)) %>% 
  select(-bill8,-sex)

monthly_m <- monthly_m %>% 
  mutate(month_year = ymd(month_year)) %>% 
  select(-bill8)

monthly_all <- monthly_all %>% 
  mutate(month_year = ymd(month_year))

```

**How many deaths occurred among females? Males? Which months had highest counts?**

```{r}
#female death count
monthly_f %>% 
  summarize(n_deaths = sum(Deaths, na.rm = TRUE))  
#male death count
monthly_m %>% 
  summarize(n_deaths = sum(Deaths, na.rm = TRUE))  
#highest counts female
monthly_f %>% 
  arrange(desc(Deaths)) %>% 
  select(month_year, Deaths) %>% 
  slice(1:3) 
#highest count male
monthly_m %>% 
  arrange(desc(Deaths)) %>% 
  select(month_year, Deaths) %>% 
  slice(1:3) 
#descriptive stats
monthly_all %>%
  group_by(sex) %>%
  summarise(
    n_months      = n(),
    total_deaths  = sum(Deaths, na.rm = TRUE),
    monthly_mean  = mean(Deaths, na.rm = TRUE),
    monthly_sd    = sd(Deaths, na.rm = TRUE),
    .groups = "drop"
  )
```
**Plotting both** 
```{r}
monthly_all <- monthly_all %>% 
   mutate(month_year = as.Date(month_year))

ggplot(monthly_all) +
  geom_line(aes(x = month_year, y = Deaths, color = sex)) +
  geom_vline(xintercept = vline_date, linetype = "dashed", color = "red") +
  annotate("text",
           x = vline_date,
           y = 45,
           label = "June, 2021",
           angle = 90,
           vjust = -0.5,
           color = "red") +
  theme_minimal()
```

## 3. All-age Assault Mortality

We will investigate assault mortality among all ages, not just reproductive. 

```{r, include = FALSE}
m_all_age <- read.csv("../CleanData/all_age_m_assault.csv")
f_all_age <- read.csv("../CleanData/all_age_f_assault.csv") 

m_all_age <- m_all_age %>% 
   mutate(month_year = as.Date(month_year))

f_all_age <- f_all_age %>% 
   mutate(month_year = as.Date(month_year))
```

```{r}
m_all_age %>%
  summarise(
    n_months      = n(),
    total_deaths  = sum(Deaths, na.rm = TRUE),
    monthly_mean  = mean(Deaths, na.rm = TRUE),
    monthly_sd    = sd(Deaths, na.rm = TRUE),
    .groups = "drop"
  )

f_all_age %>%
  summarise(
    n_months      = n(),
    total_deaths  = sum(Deaths, na.rm = TRUE),
    monthly_mean  = mean(Deaths, na.rm = TRUE),
    monthly_sd    = sd(Deaths, na.rm = TRUE),
    .groups = "drop"
  )

``` 
**Plotting both** 
```{r}
ggplot() +
  geom_line(data = f_all_age,
            aes(x = month_year, y = Deaths),
            color = "hotpink") +
  geom_line(data = m_all_age,
            aes(x = month_year, y = Deaths),
            color = "lightblue3") +
  labs(x = "Date",
       y = "Deaths") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  geom_vline(xintercept = vline_date, linetype = "dashed", color = "red") +
  annotate("text",
           x = vline_date,
           y = 65,
           label = "June, 2021",
           angle = 90,
           vjust = -0.5,
           color = "red") +
  theme_minimal()+ 
  ggtitle("All Age Assault Deaths", 
          subtitle= "Male vs. Female")
```

## 4. Racial breakdown 

```{r, include = FALSE}
racial_minus_covid <- read.csv("../CleanData/racial_minus_covid.csv") 

racial_minus_covid <- racial_minus_covid %>% 
  mutate(month_year = as.Date(month_year))
``` 

**Summary Stats by Race** 

```{r}
racial_minus_covid %>%
  group_by(race_eth) %>%
  summarise(
    n_months      = n(),
    total_deaths  = sum(Deaths, na.rm = TRUE),
    monthly_mean  = mean(Deaths, na.rm = TRUE),
    monthly_sd    = sd(Deaths, na.rm = TRUE),
    .groups = "drop"
  )

``` 

**Highest Death Count Months** 
```{r}
racial_minus_covid %>%
  arrange(desc(Deaths)) %>% 
  group_by(race_eth) %>% 
  slice(1:3) 
```


**Plotting racial trends**

```{r}
ggplot(racial_minus_covid,
       aes(x = month_year, y = Deaths, color = race_eth)) +
  geom_line() +
  labs(x = "Date",
       y = "Deaths",
       color = "Race/Ethnicity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_vline(xintercept = vline_date, linetype = "dashed", color = "red") +
  annotate("text",
           x = vline_date,
           y = 45,
           label = "June, 2021",
           angle = 90,
           vjust = -0.5,
           color = "red") +
  ggtitle("All-Cause Mortality (Excluding COVID-19) by Race/Ethnicity")

```


# Investigating Models

Next we are going to plot the forcing value (time) and the outcome (death count)
```{r}
#looking at range of dates 
range(monthly_f$time)

ggplot()+
  geom_polygon(aes(c(45,70,70,45),
                   c(Inf, Inf, -Inf, -Inf)),
               fill = "cyan", alpha = 0.2)+
  geom_line(data = monthly_f, aes(time, Deaths))+
  labs(x = "Month in study",
       y = "Number of deaths by assault")

```

We want to investigate the death rate before and after the ban was implemented. To start, we are just going to use a simple poisson model.

```{r poisson_ITS}
mod1 <- glm(Deaths ~ time , 
                      data = monthly_f, 
                       family = poisson(link = "log"))
summary(mod1) 

#introducing bill8
mod2 <- glm(
  Deaths ~ time + bill8,
  family = poisson(link = "log"),
  data = monthly_f
)

summary(mod2) 

#extracting coefficient terms 
mod2 %>% 
  tidy()

#getting estimate of relative rate for group after bill vs. before

mod2 %>%
  tidy(conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term == "bill8TRUE") %>%
  select(estimate, conf.low, conf.high)

#how to plot the interrupted ts 

#saving augmented data from model
mod2_data <- augment(mod2, type.predict = "response") %>%
  mutate(
    observed = Deaths,
    fitted   = .fitted
  ) 
intervention_time <- min(monthly_f$time[monthly_f$bill8])

#ggplotting
ggplot(mod2_data, aes(x = time)) + 
  #post-bill shading
   geom_rect(
    aes(
      xmin = intervention_time,
      xmax = max(time),
      ymin = -Inf,
      ymax = Inf
    ),
    fill = "cyan",
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  #observed death counts
   geom_line(aes(y = observed), alpha = 0.7) + 
  #fit ITS line 
  geom_line(aes(y = fitted), color = "red", linewidth = 1) +

  labs(
    x = "Time (months)",
    y = "Monthly deaths",
    title = "Interrupted Time Series Analysis",
    subtitle = "Poisson model with level change after Bill 8"
  ) +
  theme_minimal()
  
```

Looking at that first poisson model, there is no significance between groups. We expected that, however, as the poisson isn't the most appropriate selection and our data has seasonality trends. We are now going to look into making this a nonlinear model. 

```{r seasonality}
#making month_year a date var 
monthly_f <- monthly_f %>%
  mutate(
    month_year = as.Date(month_year),
    month = month(month_year)
  ) 

#post-intervention slope term
intervention_time <- min(monthly_f$time[monthly_f$bill8])

monthly_f <- monthly_f %>%
  mutate(
    time_after = if_else(bill8, time - intervention_time + 1, 0)
  )

monthly_f %>% 
  group_by(time_after) %>% 
  slice(1:3) 


mod3 <- glm(
  Deaths ~ bill8 + time +
    harmonic(month, 2, 12),
  data = monthly_f,
  family = quasipoisson(link = "log")
)


#plotting with seasonality
plot_data <- augment(mod3, type.predict = "response") %>%
  mutate(
    observed = Deaths,
    fitted = .fitted
  )

ggplot(plot_data, aes(x = time)) +

  geom_rect(
    aes(
      xmin = intervention_time,
      xmax = max(time),
      ymin = -Inf,
      ymax = Inf
    ),
    fill = "cyan",
    alpha = 0.2,
    inherit.aes = FALSE
  ) +

  geom_line(aes(y = observed), alpha = 0.6) +
  geom_line(aes(y = fitted), color = "red", linewidth = 1) +

  labs(
    x = "Time (months)",
    y = "Monthly deaths",
    title = "Interrupted Time Series with Seasonality Adjustment",
    subtitle = "Quasi-Poisson model with harmonic terms"
  ) +
  theme_minimal()


``` 


This provides us some interesting trends, but what we are really interested is utilizing an interrupted-ARIMA model to identify the significance of our outliers. We are going to proceed with this analysis in the rmd document called `/Scripts/Analysis`.  



I am next going to replicate Figure 1 from the manuscript: 
```{r figure_1 ,fig.width = 11, fig.height = 7}

# Colors
colors <- c("Female" = "#A50F15", "Male" = "#555555")  

# Separate female & male
female <- monthly_all %>% filter(sex == "Female")
male   <- monthly_all %>% filter(sex == "Male")

# Scale factor for secondary axis
scale_factor <- max(female$Deaths, na.rm = TRUE) / max(male$Deaths, na.rm = TRUE)

# Vertical line time
vline_date <- as.Date("2021-09-01")

ggplot() +
  # Female line
  geom_line(data = female, aes(x = month_year, y = Deaths, color = "Female"), size = 1) +
  # Male line (scaled for secondary axis)
  geom_line(data = male, aes(x = month_year, y = Deaths * scale_factor, color = "Male"), size = 1) +
  
  # Vertical dashed line with label
  geom_vline(xintercept = vline_date, linetype = "dashed", color = "black") +
  annotate("text",
           x = vline_date,
           y = max(female$Deaths, na.rm = TRUE) + 5, # slightly above top
           label = "Sep, 2021",
           angle = 90,
           vjust = -0.5) +

  # Scale axes
  scale_y_continuous(
    name = "Monthly count among women",
    limits = c(0, max(female$Deaths, na.rm = TRUE) + 10),  # expand y-axis for better readability of date
    sec.axis = sec_axis(~./scale_factor,
                        name = "Monthly count among men")
  ) +
  
  # X-axis formatting
  scale_x_date(date_labels = "%Ym%m", date_breaks = "6 months") +
  
  # Line colors
  scale_color_manual(values = colors) +
  
  # Theme tweaks for colored axis titles
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.key.width = unit(2, "lines"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.y.left = element_text(color = colors["Female"], size = 12, face = "bold"),
    axis.title.y.right = element_text(color = colors["Male"], size = 12, face = "bold")
  ) +
  labs(x = "Time (monthly)", color = NULL)


```

