---
title: "JuneInterruption"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
---

```{r setup}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(broom)
library(viridis)
library(tsModel)
library(scales)
library(grid) 
library(tseries) 
library(tsibble) 
library(fable)
library(feasts) 
library(forecast)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
source("functions.R")

``` 

# All-Cause
```{r import, include = FALSE}
monthly_f <- read.csv("../CleanData/all_cause_repro_f.csv") 
monthly_m <- read.csv("../CleanData/all_cause_repro_m.csv")

monthly_f <- monthly_f %>% 
  mutate(month_year = ymd(month_year))

monthly_m <- monthly_m %>% 
  mutate(month_year = ymd(month_year))

monthly_f.ts <- monthly_f %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_f.ts) 

monthly_m.ts <- monthly_m %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_m.ts)

``` 

```{r ACF_PACF, include = FALSE}
monthly_f.ts  |> #female
  gg_tsdisplay(y = Deaths, plot_type = "partial")

monthly_m.ts  |> #male
  gg_tsdisplay(y = Deaths, plot_type = "partial")
```


```{r, echo = FALSE}
# Create variable representing step and ramp changes
intervention_month <- monthly_f.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_f.ts<- monthly_f.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  ) 

intervention_month <- monthly_m.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_m.ts <- monthly_m.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  )

#View the step and ramp variables
monthly_f.ts %>%  #female
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Female",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 


monthly_m.ts %>% #male
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Male",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 

```
## ARIMA model 

The Fable algorithm will dictate the model parameters. 
Next, we plot the residuals, the ACF plot, and a histogram to ensure they approximate a white noise series. 

```{r, results='hide', message = FALSE, warning = FALSE}
# Use automated algorithm to identify p/q parameters
# Specify first difference = 1 and seasonal difference = 1
f_arima <- monthly_f.ts %>%  #female
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
m_arima <- monthly_m.ts %>%  #male
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
``` 


```{r}
#Plotting model residuals
f_arima  %>% 
  gg_tsresiduals() 

m_arima  %>% 
  gg_tsresiduals()
``` 

Here we use the Ljung-Box test, which models if the residuals are still autocorrelated. 
Null: residuals are independent (white noise)
Alternative: residuals show remaining autocorrelation 
```{r ljung-box2}
#Ljung-Box test
augment(f_arima ) %>%  
  features(.resid, ljung_box) 

augment(m_arima ) %>%  
  features(.resid, ljung_box)

```

With the model assumptions satisfied, we can take a look at the model estimates: 
```{r fable_arima2}
report(m_arima)
report(f_arima)

```  


```{r}
#Modeling counterfactual male
m_counterfactual <- monthly_m.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(m_counterfactual) 

#Modeling counterfactual female
f_counterfactual <- monthly_f.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(f_counterfactual)

#forecast into post-intervention period  male
fc_m_counterfactual <- m_counterfactual %>%
  forecast(h = 28) 

 autoplot(fc_m_counterfactual, monthly_m.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Male Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) 
 
 #forecast into post-intervention period male
fc_f_counterfactual <- f_counterfactual %>%
  forecast(h = 28) 

 autoplot(fc_f_counterfactual, monthly_f.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Female Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

``` 
## Sensitivity 

**Step 1: Fit ARIMA using pre-intervention data only**  

Subsetting pre-intervention data:  

```{r}
pre_data_f <- monthly_f.ts %>%
  filter(month_year < as.Date("2021-06-01")) 

pre_data_m <- monthly_m.ts %>%
  filter(month_year < as.Date("2021-06-01")) 
```

Selecting ARIMA from pre-period data only: 

```{r arima_specs2, results='hide', message = FALSE, warning = FALSE}
m_pre <- pre_data_m %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))

f_pre <- pre_data_f %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))
``` 

```{r}
report(m_pre)
report(f_pre)
```

**Step 2: Refit to full data** 

```{r}
m_sensitivity <- monthly_m.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,0) + PDQ(1,0,0),
      stepwise = FALSE
    )
  ) 

f_sensitivity <- monthly_f.ts %>% 
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(1,1,3),
      stepwise = FALSE
    )
  ) 

```

**Step 3: Comparing estimates** 
```{r} 
#male
report(m_arima)
report(m_sensitivity) 

#female 
report(f_arima)
report(f_sensitivity)
``` 

# Assault COD 

```{r import, include = FALSE}
monthly_f <- read.csv("../CleanData/monthly_f.csv") 
monthly_m <- read.csv("../CleanData/monthly_m.csv")

monthly_f <- monthly_f %>% 
  mutate(month_year = ymd(month_year))

monthly_m <- monthly_m %>% 
  mutate(month_year = ymd(month_year))

monthly_f.ts <- monthly_f %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_f.ts) 

monthly_m.ts <- monthly_m %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_m.ts)

``` 


```{r ACF_PACF, include = FALSE}
monthly_f.ts  |> #female
  gg_tsdisplay(y = Deaths, plot_type = "partial")

monthly_m.ts  |> #male
  gg_tsdisplay(y = Deaths, plot_type = "partial")
```


```{r}
# Create variable representing step and ramp changes
intervention_month <- monthly_f.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_f.ts<- monthly_f.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  ) 

intervention_month <- monthly_m.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_m.ts <- monthly_m.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  )

#View the step and ramp variables
monthly_f.ts %>%  #female
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Female",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 


monthly_m.ts %>% #male
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Male",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 

```
## ARIMA model 

The Fable algorithm will dictate the model parameters. 
Next, we plot the residuals, the ACF plot, and a histogram to ensure they approximate a white noise series. 

```{r, results='hide', message = FALSE, warning = FALSE}
# Use automated algorithm to identify p/q parameters
# Specify first difference = 1 and seasonal difference = 1
f_arima <- monthly_f.ts %>%  #female
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
m_arima <- monthly_m.ts %>%  #male
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
``` 


```{r}
#Plotting model residuals
f_arima  %>% 
  gg_tsresiduals() 

m_arima  %>% 
  gg_tsresiduals()
``` 

Here we use the Ljung-Box test, which models if the residuals are still autocorrelated. 
Null: residuals are independent (white noise)
Alternative: residuals show remaining autocorrelation 
```{r ljung-box2}
#Ljung-Box test
augment(f_arima ) %>%  
  features(.resid, ljung_box) 

augment(m_arima ) %>%  
  features(.resid, ljung_box)

```

With the model assumptions satisfied, we can take a look at the model estimates: 
```{r fable_arima2}
report(m_arima)
report(f_arima)

```  


```{r}
#Modeling counterfactual male
m_counterfactual <- monthly_m.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(m_counterfactual) 

#Modeling counterfactual female
f_counterfactual <- monthly_f.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(f_counterfactual)

#forecast into post-intervention period  male
fc_m_counterfactual <- m_counterfactual %>%
  forecast(h = 28) 

 autoplot(fc_m_counterfactual, monthly_m.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Male Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) 
 
 #forecast into post-intervention period male
fc_f_counterfactual <- f_counterfactual %>%
  forecast(h = 28) 

 autoplot(fc_f_counterfactual, monthly_f.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Female Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```  


## Sensitivity 

**Step 1: Fit ARIMA using pre-intervention data only**  

Subsetting pre-intervention data:  

```{r}
pre_data_f <- monthly_f.ts %>%
  filter(month_year < as.Date("2021-06-01")) 

pre_data_m <- monthly_m.ts %>%
  filter(month_year < as.Date("2021-06-01")) 
```

Selecting ARIMA from pre-period data only: 

```{r arima_specs2, results='hide', message = FALSE, warning = FALSE}
m_pre <- pre_data_m %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))

f_pre <- pre_data_f %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))
``` 

```{r}
report(m_pre)
report(f_pre)
```

**Step 2: Refit to full data** 

```{r}
m_sensitivity <- monthly_m.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,1),
      stepwise = FALSE
    )
  ) 

f_sensitivity <- monthly_f.ts %>% 
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(4,1,0),
      stepwise = FALSE
    )
  ) 

```

**Step 3: Comparing estimates** 
```{r} 
#male
report(m_arima)
report(m_sensitivity) 

#female 
report(f_arima)
report(f_sensitivity)
``` 
# All ages 

```{r import, include = FALSE}
monthly_f <- read.csv("../CleanData/all_age_f_assault.csv") 
monthly_m <- read.csv("../CleanData/all_age_m_assault.csv")

monthly_f <- monthly_f %>% 
  mutate(month_year = ymd(month_year))

monthly_m <- monthly_m %>% 
  mutate(month_year = ymd(month_year))

monthly_f.ts <- monthly_f %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_f.ts) 

monthly_m.ts <- monthly_m %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_m.ts)

``` 

```{r ACF_PACF, include = FALSE}
monthly_f.ts  |> #female
  gg_tsdisplay(y = Deaths, plot_type = "partial")

monthly_m.ts  |> #male
  gg_tsdisplay(y = Deaths, plot_type = "partial")
``` 

```{r}
# Create variable representing step and ramp changes
intervention_month <- monthly_f.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_f.ts<- monthly_f.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  ) 

intervention_month <- monthly_m.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_m.ts <- monthly_m.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  )

#View the step and ramp variables
monthly_f.ts %>%  #female
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Female",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 


monthly_m.ts %>% #male
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Male",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 

```


## ARIMA model 

The Fable algorithm will dictate the model parameters. 
Next, we plot the residuals, the ACF plot, and a histogram to ensure they approximate a white noise series. 

```{r, results='hide', message = FALSE, warning = FALSE}
# Use automated algorithm to identify p/q parameters
# Specify first difference = 1 and seasonal difference = 1
f_arima <- monthly_f.ts %>%  #female
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
m_arima <- monthly_m.ts %>%  #male
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
``` 


```{r}
#Plotting model residuals
f_arima  %>% 
  gg_tsresiduals() 

m_arima  %>% 
  gg_tsresiduals()
```  

Here we use the Ljung-Box test, which models if the residuals are still autocorrelated. 
Null: residuals are independent (white noise)
Alternative: residuals show remaining autocorrelation 
```{r ljung-box2}
#Ljung-Box test
augment(f_arima ) %>%  
  features(.resid, ljung_box) 

augment(m_arima ) %>%  
  features(.resid, ljung_box)

```

With the model assumptions satisfied, we can take a look at the model estimates: 
```{r fable_arima2}
report(m_arima)
report(f_arima)

```   

```{r}
#Modeling counterfactual male
m_counterfactual <- monthly_m.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(m_counterfactual) 

#Modeling counterfactual female
f_counterfactual <- monthly_f.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(f_counterfactual)

#forecast into post-intervention period  male
fc_m_counterfactual <- m_counterfactual %>%
  forecast(h = 28) 

 autoplot(fc_m_counterfactual, monthly_m.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Male Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) 
 
 #forecast into post-intervention period male
fc_f_counterfactual <- f_counterfactual %>%
  forecast(h = 28) 

 autoplot(fc_f_counterfactual, monthly_f.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Female Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```  

## Sensitivity 

**Step 1: Fit ARIMA using pre-intervention data only**  

Subsetting pre-intervention data:  

```{r}
pre_data_f <- monthly_f.ts %>%
  filter(month_year < as.Date("2021-06-01")) 

pre_data_m <- monthly_m.ts %>%
  filter(month_year < as.Date("2021-06-01")) 
```

Selecting ARIMA from pre-period data only: 

```{r arima_specs2, results='hide', message = FALSE, warning = FALSE}
m_pre <- pre_data_m %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))

f_pre <- pre_data_f %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))
``` 

```{r}
report(m_pre)
report(f_pre)
```

**Step 2: Refit to full data** 

```{r}
m_sensitivity <- monthly_m.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,1),
      stepwise = FALSE
    )
  ) 

f_sensitivity <- monthly_f.ts %>% 
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,1),
      stepwise = FALSE
    )
  ) 

```

**Step 3: Comparing estimates** 
```{r} 
#male
report(m_arima)
report(m_sensitivity) 

#female 
report(f_arima)
report(f_sensitivity)
``` 
