---
title: "Interrupted Time Series ARIMA Analysis Bill 8"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
    code_folding: hide
--- 


```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(broom)
library(viridis)
library(tsModel)
library(scales)
library(grid) 
library(tseries) 
library(tsibble) 
library(fable)
library(feasts) 
library(forecast)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE) 
source("functions.R")

``` 

# Introduction 

Rationale for June interruption: May 29th the bill got signed in, was in the media, want to capture effects immediately after and also preserve the spike we notice (which would be factored into the forecast if we made the interruption point September)


# Results 

# Methods (Expandable)
  - Data preparation
  - Intervention variables
  - ARIMA specification

# Model diagnostics (Expandable)

# Sensitivity analysis (Expandable)


# All-Cause
```{r, include = FALSE}
monthly_f <- read.csv("../CleanData/all_cause_repro_f.csv") 
monthly_m <- read.csv("../CleanData/all_cause_repro_m.csv")

monthly_f <- monthly_f %>% 
  mutate(month_year = ymd(month_year))

monthly_m <- monthly_m %>% 
  mutate(month_year = ymd(month_year))

monthly_f.ts <- monthly_f %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_f.ts) 

monthly_m.ts <- monthly_m %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_m.ts)

``` 

```{r, include = FALSE}
monthly_f.ts  |> #female
  gg_tsdisplay(y = Deaths, plot_type = "partial")

monthly_m.ts  |> #male
  gg_tsdisplay(y = Deaths, plot_type = "partial")
```


```{r, echo = FALSE}
# Create variable representing step and ramp changes
intervention_month <- monthly_f.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_f.ts<- monthly_f.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  ) 

intervention_month <- monthly_m.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_m.ts <- monthly_m.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  )

#View the step and ramp variables
monthly_f.ts %>%  #female
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Female",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 


monthly_m.ts %>% #male
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Male",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 

```

## ARIMA model 

The Fable algorithm will dictate the model parameters. 
Next, we plot the residuals, the ACF plot, and a histogram to ensure they approximate a white noise series. 

```{r, results='hide', message = FALSE, warning = FALSE}
# Use automated algorithm to identify p/q parameters
# Specify first difference = 1 and seasonal difference = 1
f_arima <- monthly_f.ts %>%  #female
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
m_arima <- monthly_m.ts %>%  #male
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
``` 


```{r}
#Plotting model residuals
f_arima  %>% 
  gg_tsresiduals() 

m_arima  %>% 
  gg_tsresiduals()
``` 


```{r, include = FALSE}
#Ljung-Box test
augment(f_arima ) %>%  
  features(.resid, ljung_box) 

augment(m_arima ) %>%  
  features(.resid, ljung_box)

```

Model estimates: 
```{r}
report(m_arima)
report(f_arima)
```  

## Forecast Graph 

```{r, echo = TRUE}
#Modeling counterfactual male
m_counterfactual <- monthly_m.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(m_counterfactual) 

#Modeling counterfactual female
f_counterfactual <- monthly_f.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(f_counterfactual)

#forecast into post-intervention period  male
fc_m_counterfactual <- m_counterfactual %>%
  forecast(h = 28) 

 autoplot(fc_m_counterfactual, monthly_m.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Male All-Cause Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) 
 
 #forecast into post-intervention period male
fc_f_counterfactual <- f_counterfactual %>%
  forecast(h = 28+6) 

 autoplot(fc_f_counterfactual, monthly_f.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Female All-Cause Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

``` 

## Sensitivity Analysis

 
```{r, include = FALSE} 
#Step 1: Fit ARIMA using pre-intervention data only  
pre_data_f <- monthly_f.ts %>%
  filter(month_year < as.Date("2021-06-01")) 

pre_data_m <- monthly_m.ts %>%
  filter(month_year < as.Date("2021-06-01")) 
```


```{r, include = FALSE, message = FALSE, warning = FALSE}
m_pre <- pre_data_m %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))

f_pre <- pre_data_f %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))
``` 


```{r, include = FALSE} 
#pre-intervention
report(m_pre)
report(f_pre)
```


```{r, include = FALSE}
m_sensitivity <- monthly_m.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,0) + PDQ(1,0,0),
      stepwise = FALSE
    )
  ) 

f_sensitivity <- monthly_f.ts %>% 
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(1,1,3),
      stepwise = FALSE
    )
  ) 

```

**Comparing estimates** 
(Fable vs. Sensitivity)
```{r} 
#male
report(m_arima)
report(m_sensitivity) 

#female 
report(f_arima)
report(f_sensitivity)
``` 

# Assault COD 

```{r, include = FALSE}
monthly_f <- read.csv("../CleanData/monthly_f.csv") 
monthly_m <- read.csv("../CleanData/monthly_m.csv")

monthly_f <- monthly_f %>% 
  mutate(month_year = ymd(month_year))

monthly_m <- monthly_m %>% 
  mutate(month_year = ymd(month_year))

monthly_f.ts <- monthly_f %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_f.ts) 

monthly_m.ts <- monthly_m %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_m.ts)

``` 


```{r, include = FALSE}
monthly_f.ts  |> #female
  gg_tsdisplay(y = Deaths, plot_type = "partial")

monthly_m.ts  |> #male
  gg_tsdisplay(y = Deaths, plot_type = "partial")
```


```{r, include = FALSE}
# Create variable representing step and ramp changes
intervention_month <- monthly_f.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_f.ts<- monthly_f.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  ) 

intervention_month <- monthly_m.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_m.ts <- monthly_m.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  )

#View the step and ramp variables
monthly_f.ts %>%  #female
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Female",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 


monthly_m.ts %>% #male
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Male",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 

```

## ARIMA model 

The Fable algorithm will dictate the model parameters. 
Next, we plot the residuals, the ACF plot, and a histogram to ensure they approximate a white noise series. 

```{r, include = FALSE, message = FALSE, warning = FALSE}
# Use automated algorithm to identify p/q parameters
# Specify first difference = 1 and seasonal difference = 1
f_arima <- monthly_f.ts %>%  #female
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
m_arima <- monthly_m.ts %>%  #male
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
``` 


```{r, include = FALSE}
#Plotting model residuals
f_arima  %>% 
  gg_tsresiduals() 

m_arima  %>% 
  gg_tsresiduals()
``` 


```{r, include = FALSE}
#Ljung-Box test
augment(f_arima ) %>%  
  features(.resid, ljung_box) 

augment(m_arima ) %>%  
  features(.resid, ljung_box)

```

Model estimates: 
```{r}
report(m_arima)
report(f_arima)

```  


```{r, echo = FALSE}
#Modeling counterfactual male
m_counterfactual <- monthly_m.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(m_counterfactual) 

#Modeling counterfactual female
f_counterfactual <- monthly_f.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(f_counterfactual)

#forecast into post-intervention period  male
fc_m_counterfactual <- m_counterfactual %>%
  forecast(h = 28+6) 

 autoplot(fc_m_counterfactual, monthly_m.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Male Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) 
 
 #forecast into post-intervention period male
fc_f_counterfactual <- f_counterfactual %>%
  forecast(h = 28+6) 

 autoplot(fc_f_counterfactual, monthly_f.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual Female Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```  


## Sensitivity 

**Step 1: Fit ARIMA using pre-intervention data only**  

Subsetting pre-intervention data:  

```{r, include = FALSE}
pre_data_f <- monthly_f.ts %>%
  filter(month_year < as.Date("2021-06-01")) 

pre_data_m <- monthly_m.ts %>%
  filter(month_year < as.Date("2021-06-01")) 
```



```{r,include = FALSE, message = FALSE, warning = FALSE}
m_pre <- pre_data_m %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))

f_pre <- pre_data_f %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))
``` 

```{r,include = FALSE}
report(m_pre)
report(f_pre)
```



```{r, include = FALSE}
m_sensitivity <- monthly_m.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,1),
      stepwise = FALSE
    )
  ) 

f_sensitivity <- monthly_f.ts %>% 
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(4,1,0),
      stepwise = FALSE
    )
  ) 

```

**Step 3: Comparing estimates** 
```{r} 
#male
report(m_arima)
report(m_sensitivity) 

#female 
report(f_arima)
report(f_sensitivity)
``` 

# All ages 

```{r, include = FALSE}
monthly_f <- read.csv("../CleanData/all_age_f_assault.csv") 
monthly_m <- read.csv("../CleanData/all_age_m_assault.csv")

monthly_f <- monthly_f %>% 
  mutate(month_year = ymd(month_year))

monthly_m <- monthly_m %>% 
  mutate(month_year = ymd(month_year))

monthly_f.ts <- monthly_f %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_f.ts) 

monthly_m.ts <- monthly_m %>% 
  mutate(month_year = ymd(month_year),
         month = yearmonth(month_year)) %>% 
  as_tsibble(index = month)
  
glimpse(monthly_m.ts)

``` 

```{r, include = FALSE}
monthly_f.ts  |> #female
  gg_tsdisplay(y = Deaths, plot_type = "partial")

monthly_m.ts  |> #male
  gg_tsdisplay(y = Deaths, plot_type = "partial")
``` 

```{r,include = FALSE}
# Create variable representing step and ramp changes
intervention_month <- monthly_f.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_f.ts<- monthly_f.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  ) 

intervention_month <- monthly_m.ts %>% 
  filter(month_year == "2021-06-01") %>% 
  pull(time)

monthly_m.ts <- monthly_m.ts %>%
  mutate(
    t = row_number(),
    step = if_else(t >= intervention_month, 1, 0),
    ramp = if_else(t >= intervention_month, t - intervention_month + 1, 0)
  )

#View the step and ramp variables
monthly_f.ts %>%  #female
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Female",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 


monthly_m.ts %>% #male
  pivot_longer(
    cols = c(step, ramp), 
    names_to = "variables",
    values_to = "value"
  ) %>%
  ggplot(aes(x = month_year, y = value, color = variables)) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = as.Date("2021-06-01"),
    linetype = "dashed",
    color = "black"
  ) +
  labs(
    title = "Step and Ramp Variables - Male",
    x = "Month",
    y = "Value",
    color = ""
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_minimal() 

```


## ARIMA model 

The Fable algorithm will dictate the model parameters. 
Next, we plot the residuals, the ACF plot, and a histogram to ensure they approximate a white noise series. 

```{r, results='hide', message = FALSE, warning = FALSE}
# Use automated algorithm to identify p/q parameters
# Specify first difference = 1 and seasonal difference = 1
f_arima <- monthly_f.ts %>%  #female
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
m_arima <- monthly_m.ts %>%  #male
  model(arima = ARIMA(Deaths ~ step + ramp,
                      stepwise = FALSE, 
                      trace = TRUE)) 
``` 


```{r,include = FALSE}
#Plotting model residuals
f_arima  %>% 
  gg_tsresiduals() 

m_arima  %>% 
  gg_tsresiduals()
```  

 
```{r,include = FALSE}
#Ljung-Box test
augment(f_arima ) %>%  
  features(.resid, ljung_box) 

augment(m_arima ) %>%  
  features(.resid, ljung_box)

```

Model estimates: 
```{r}
report(m_arima)
report(f_arima)

```   

```{r}
#Modeling counterfactual male
m_counterfactual <- monthly_m.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(m_counterfactual) 

#Modeling counterfactual female
f_counterfactual <- monthly_f.ts %>%
  filter(month < make_yearmonth(2021, 6)) %>%
  model(
    arima_null = ARIMA(
      Deaths,
      stepwise = FALSE
    )
  ) 
report(f_counterfactual)

#forecast into post-intervention period  male
fc_m_counterfactual <- m_counterfactual %>%
  forecast(h = 28+6) 

 autoplot(fc_m_counterfactual, monthly_m.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual All Age Male Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) 
 
 #forecast into post-intervention period male
fc_f_counterfactual <- f_counterfactual %>%
  forecast(h = 28+6) 

 autoplot(fc_f_counterfactual, monthly_f.ts, level = 95) +
  geom_vline(
    xintercept = as.numeric(as_date(make_yearmonth(2021, 6))),
    linetype = "dashed",
    color = "grey40"
  ) +
  labs(
    title = "Observed vs Counterfactual All Age Female Assault Deaths",
    y = "Monthly Deaths",
    x = "Month"
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_minimal() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

```  

## Sensitivity 

**Step 1: Fit ARIMA using pre-intervention data only**  

Subsetting pre-intervention data:  

```{r,include = FALSE}
pre_data_f <- monthly_f.ts %>%
  filter(month_year < as.Date("2021-06-01")) 

pre_data_m <- monthly_m.ts %>%
  filter(month_year < as.Date("2021-06-01")) 
```


```{r, results='hide', message = FALSE, warning = FALSE}
m_pre <- pre_data_m %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))

f_pre <- pre_data_f %>%
  model(arima = ARIMA(Deaths, stepwise = FALSE))
``` 

```{r,include = FALSE}
report(m_pre)
report(f_pre)
```

**Step 2: Refit to full data** 

```{r,include = FALSE}
m_sensitivity <- monthly_m.ts %>%
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,1),
      stepwise = FALSE
    )
  ) 

f_sensitivity <- monthly_f.ts %>% 
  model(
    arima = ARIMA(
      Deaths ~ step + ramp + pdq(0,1,1),
      stepwise = FALSE
    )
  ) 

```

**Step 3: Comparing estimates** 
```{r} 
#male
report(m_arima)
report(m_sensitivity) 

#female 
report(f_arima)
report(f_sensitivity)
``` 
